Individual Variation in Animal Movement and Effects on Seed Dispersal Distance

### 2020-10-14 paper/paper.Rmd
word output
progress

abstract: |
  Seed dispersal is a fundamental process in plant ecology and due to its high importance in the ecosystem, various modelling approaches have been developed to understand seed movement and dispersal mechanisms. In particular, long-distance seed dispersal (LDD) plays a significant role in determining a population's genetic diversity and range expansion. Animal-mediated seed dispersal is a key component of LDD events, as frugivores can travel long distances and transport seeds with them. In this study we seek to understand the implications of individual variation in animal movement and how these can impact estimates of long distance seed dispersal. We apply these methods to empirical data sets that contain locations for an aracari, *Pteroglossus pluricintus*, one of the primary frugivores for the Amazonian canopy tree *Virola flexuosa*. We combine animal movement data and gut retention time to simulate *Virola* seed dispersal. This approach showed that there is significant variation in individual aracari movement patterns which directly influence the estimated number of long-distance dispersal events for *Virola* seeds. We found that models that ignore this underlying variation in individual animal movement, underestimate the percentage of long-distance seed dispersal events and maximum seed dispersal distances.
output: html_document
bibliography: mybibfile.bib
---

@@

In order to analyze the effect of differences in animal movement over final dispersal distances of foraged seeds we developed a spatially-explicit individual-based model in a homogeneous landscape that incorporated animal movement and seed gut retention time (GRT). Our landscape consisted of a single source tree at the origin, our simulation time steps were based on one minute intervals, and each simulation started at the source tree (0,0), where each individual bird was given five seeds and was allowed to move freely in the landscape until the simulation ended once all seeds were dropped. Each seed's gut retention time was determined at the start of each simulation run, being randomly sampled from a Gamma distribution (shape = 4, scale = 5, shift = 8) [@morales_effects_2006] so that the average gut retention time was 28 minutes[@holbrook_dispersal_2009; @holbrook_using_2007]. For each simulation run, the total simulation time was determined by the largest gut retention time sampled for the five seeds, therefore, once the last seed was dropped, the simulation ended. From the simulation, we calculated the Euclidean distance from the source tree to each of the dropped seeds to determine the seed dispersal distance. We simulated animal movement using a memoryless process, a random walk with a uniformly distributed angular direction and a random step size chosen from a probability distribution. The simulation proceeded as follows: an individual bird starts the simulation at time zero and location zero zero. That bird is given five different seeds, with gut retention time (GRT_I), where GRT is gamma distributed. The animal is allowed to move freely across the landscape where at each time point T_i, a movement angle is sampled ~uniform(360) (think of greek letter) and a step size is randomly selected d_i, where D ~ f(x), where f(x) is the probability density function of one of the models we've selected. The movement process is repeated and at T_i == GRT_i a seed is dropped at the bird's location (x_i, y_i) and therefore that seed's final locations is (x_i, y_i). The simualation ends at T_i = max(GRT_1:5), and seed dispersal distances are calculated from the origin (0,0) to each seed's location (x_i, y_i).

@@

Our main hypothesis is about the incoporation of variance in animal movement at different organizational levels, meaning how are the differences in individual animal movement affecting population level estimates of seed dispersal. How does the underlying variation in individual animal movement impact our estimates of long distance seed dispersal. We explored the effects of individual differences in movement by modifying our base simulation, in which the probability density function used to model animal movement was different between populations or a subcategory. In simulations at the population level, we used a single pdf to sample step sizes from, while for individual and family level simulations we used multiple pdfs to sample step sizes from, where each pdf was associated to a specific individual or family group. The pdf for the population level is f(x, sigma), where sigma is the parameters for population, and fx(x, sigma_vector), is for the individual or family level and sigma vector represents a vector of parameters for the pdf, where each parameter corresponds to a different individual or family group. We used four different pdfs for our simulations based on used probability distributions for sampling the animal movement step size including an exponential distribution [@jones_closing_2017], gamma distribution [citation], Weibull distribution[citation], and lognormal distribution [@levey_modelling_2008]. These four distributions are leptokurtic, and they vary in the fatness of their tails, thus allowing for different levels of variation in step size. We also performed a mixed distribution simulation at the individual and population levels, where each individual or family group would not only have their specific set of parameter estimates, but we also allowed for variaiton in the structure of the pdf. Here I should write out how it looks mathematically.

## Data collection/manipulation

We focused our study on the many-banded aracari (*Pterglossus pluricinctus*) and its role as a frugivore of the *Virola flexuosa* tree. A subset of animal movement data collected in a previous study [@holbrook_home_2011] was selected and used towards our models and simulations. In this previous study, Aracari home ranges, average movement distances, patterns and movement rates were estimated for tracked birds over a period of four years, from 2001 to 2005, in the Ecuadorian Amazon rainforest. Further details on field methods can be found in Holbrook [-@holbrook_home_2011].  We focused on the data from the many-banded aracari, *Pteroglossus pluricinctus*, because a higher number of recorded locations were available. In addition to this, the home range of this species was found to be the largest for all the species studied, and therefore it was better suited to study long-distance seed dispersal events.

### Animal movement probability distributions

In this study, we consider population level when we consider all the tracking data together as a whole. For population level analysis we make the assumption that there is no individual variation in movement rates and therefore a single movement rate model can describe all individuals in that population. For the case of individual level analysis, we consider individual variation by analyzing the movement rates separately and fitting a probability distribution model to the tracking data for each individual. Within this framework we consider that animal movement rates (number of meters moved per minute in each individual movement bout) can be described with four different probability distribution models: an exponential distribution [@jones_closing_2017], a gamma distribution, a weibull distribution [@morales_effects_2006], and a lognormal distribution [@levey_modelling_2008]. (Comment: where do you think I should include a description of these different distributions?) The fit of each model to the data was analyzed visually via qqplots (See supplementary information).

At the population level, we compared the fit between the four distribution models using different information criteria (AIC, AICc, and BIC). At the individual level we considered five different models, four of these models corresponded to each of the four probability distributions. For the first four models, we assumed that one probability distribution could accurately describe the movement rates for each individual by changing the distribution's parameters. A fith alternative model considered instead the best fitting model for each individual and calculated the overall information criteria as a combination of the best fitting model for each individual. For each of these indivual level models, information criteria were calculated based on the number of parameters, sample size and negative log likelihood in the following way

@@

### Seed dispersal simulation

To understand the impact of individual animal movement rates over seed dispersal distances, we built a simulation model which took into account animal movement and seed gut retention time (GRT). Each simulation run consisted of an individual bird who was given five seeds at the start of the simulation. Each seed's gut retention time was randomly sampled from a Gamma distribution based on previously collected data [@holbrook_home_2011] and out of the five seeds the largest GRT determined the simulation time for each run. The simulated bird would start at point (0,0) and move across the landscape in a random walk, with angle of movement and speed determined by one minute time steps. At each minute time step, an angle of movement would be sampled from a uniform distribution and a movement rate sampled from one of movement rate models (exponential, gamma, lognormal or weibull). Seeds would get dropped at the bird's position at the time the simulation reached each seed's GRT and the simulation run would end once the bird dropped all five seeds. At the population level, we had four different simulations, each corresponding to one of the probability distributions being tested for animal movement rates. We performed 12,000 individual simulation runs with each distribution and estimated seed dispersal distances for each of the seeds. In the case of individual level simulations, we had 1000 simulation runs for each individual (twelve individuals, with a total of 12,000 simulation runs) and we did this with each of the four distribution models. Finally, we considered a (heterogeneous/mixed? need to check with JMP on terminology) model where we selected the best fitting movement rate model for each individual and used it to parameterize the simulation. For these mixed models, the probability distribution used to sample movement rates would be the best fitting distribution for each individual's movement rates (the target distribution as determined by AIC or BIC), and not a common distribution with different parameters for each individual (as in the previous individual level simulation runs).  Whether at population level or individual level, the output of the simulation runs for each model is the dispersal distance for 60,000 seeds. We performed summary statistics on seed dispersal distances for each of the models

  - What is the overall goal of this simulation?
  - What entails one simulation run
  - Overall description of how the simulation works
  - Variables that go into the simulation
  - Sources I used to reference: jones
2. Parameterization and source data
  - Data description
  - Gut retention time comes from previous literature and averages
  - Calculations at the different organizational levels
3. Evaluating fits?
4. Statistics associated to the simulations

@@

```{r}
popIC_table <- readRDS(file = "output/popIC_table.RDS")

popIC_table %>% 
  knitr::kable()
```

@@

```{r}
indIC_table <- readRDS(file = "output/indIC_table.RDS")

indIC_table %>% 
  knitr::kable()
```

@@

```{r}
AIC_table <- readRDS("manuscript/AIC_table.RDS")

AIC_table %>%
  transmute(distribution,
          aic.pop.ind = AIC.pop - AIC.ind,
         aicc.pop.ind = AICc.pop - AICc.ind,
         bic.pop.ind = BIC.pop-BIC.ind) %>% 
  knitr::kable()
```

@@

```{r}
allIC_table <- readRDS(file = "output/allIC_table.RDS")

head(allIC_table) %>% 
  knitr::kable()
```

@@

```{r}

get_all <- readRDS("output/get_all.RDS")

get_all %>% 
  ggplot(., aes(x = model, y = dispersal, color = type)) +
  geom_boxplot() +
  theme_bw()
```

Summarize some values for these distances. Is this table necessary? or how to reduce it? Do I need to make it smaller?

```{r}
get_all %>% 
  mutate(ldd = ifelse(dispersal >= 500, 1, 0),
         movement_model = paste(str_replace(model, "_move", ""), 
                                 str_sub(type, start=1, end=3), sep = "_")) -> move_model
move_model %>% 
  group_by(movement_model) %>% 
  summarise(LDD = sum(ldd),
            prcnt_ldd = 100*mean(ldd),
            avg = mean(dispersal),
            median = median(dispersal),
            max = max(dispersal)) %>% 
  arrange(prcnt_ldd) %>% 
  knitr::kable()
```

@@

>>>>

abstract: |
  Seed dispersal is a fundamental process in plant ecology and due to its high importance in the ecosystem, various modelling approaches have been developed to understand seed movement and dispersal mechanisms. In particular, long-distance seed dispersal (LDD) plays a significant role in determining a population's genetic diversity and range expansion. Animal-mediated seed dispersal is a key component of LDD events, as frugivores can travel long distances and transport seeds with them. In this study we seek to understand the implications of individual variation in animal movement and how these can impact estimates of long distance seed dispersal. We apply these methods to empirical data sets that contain locations for an aracari, *Pteroglossus pluricintus*, one of the primary frugivores for the Amazonian canopy tree *Virola flexuosa*. We combine animal movement data and gut retention time to simulate *Virola* seed dispersal. This approach showed that there is significant variation in individual aracari movement patterns which directly influence the estimated number of long-distance dispersal events for *Virola* seeds. We found that models that ignore this underlying variation in individual animal movement, underestimate the percentage of long-distance seed dispersal events and maximum seed dispersal distances.
output: word_document
bibliography: mybibfile.bib
---

@@

In order to analyze the effect of differences in animal movement over final dispersal distances of foraged seeds we developed a spatially-explicit individual-based model in a homogeneous landscape that incorporated animal movement and seed gut retention time (GRT). Our landscape consisted of a single source tree at the origin, our simulation time steps were based on one minute intervals, and each simulation started at the source tree (0,0), where each individual bird was given five seeds and was allowed to move freely in the landscape until the simulation ended once all seeds were dropped. Each seed's gut retention time was determined at the start of each simulation run, being randomly sampled from a Gamma distribution (shape = 4, scale = 5, shift = 8) [@morales_effects_2006] so that the average gut retention time was 28 minutes[@holbrook_dispersal_2009; @holbrook_using_2007]. For each simulation run, the total simulation time was determined by the largest gut retention time sampled for the five seeds, therefore, once the last seed was dropped, the simulation ended. From the simulation, we calculated the Euclidean distance from the source tree to each of the dropped seeds to determine the seed dispersal distance. We simulated animal movement using a memoryless process, a random walk with a uniformly distributed angular direction and a random step size chosen from a probability distribution.

Caption for figure 1. Simulation example. The simulation proceeded as follows: an individual bird starts the simulation at time zero and location zero zero. That bird is given five different seeds, with gut retention time (GRT_I), where GRT is gamma distributed. The animal is allowed to move freely across the landscape where at each time point T_i, a movement angle is sampled ~uniform(360) (think of greek letter) and a step size is randomly selected d_i, where D ~ f(x), where f(x) is the probability density function of one of the models we've selected. The movement process is repeated and at T_i == GRT_i a seed is dropped at the bird's location (x_i, y_i) and therefore that seed's final locations is (x_i, y_i). The simualation ends at T_i = max(GRT_1:5), and seed dispersal distances are calculated from the origin (0,0) to each seed's location (x_i, y_i).

@@

Our main hypothesis is about the incoporation of variance in animal movement at different organizational levels, meaning how are the differences in individual animal movement affecting population level estimates of seed dispersal. How does the underlying variation in individual animal movement impact our estimates of long distance seed dispersal. We explored the effects of individual differences in movement by modifying our base simulation, described above, in which the probability density function used to sample step sizes was different between population, individual, or family level simulations. What we consider as population level simulations are those where all birds share the same probability density to sample step sizes from, while for individual and family level simulations we use the same overall pdf, but allow for variation in the parameters specified for such function, where each set of parameters corresponds to a specific individual or family group. We also performed simulations in which we allowed for variation not only of parameters, but also of the overall structure of the pdf for sampling step sizes, this means that each individual or family group was allowed to have it's own pdf and its own set of parameters (Box 1 or table 1 with description of models, either the movement or simulation). We used four different probability density functions for sampling the animal movement step size, these included an exponential distribution [@jones_closing_2017], gamma distribution [citation], Weibull distribution[citation], and lognormal distribution [@levey_modelling_2008]. These four distributions are leptokurtic, and they vary in the fatness of their tails, thus allowing for different levels of variation in step size. **should add number of runs? or mention that that is in appendix B**

# Parameterization

We focused our study on the many-banded aracari (*Pterglossus pluricinctus*) and its role as a frugivore of the *Virola flexuosa* tree. A subset of animal movement data collected in a previous study [@holbrook_home_2011] was selected and used towards our models and simulations. In this previous study, Aracari home ranges, average movement distances, patterns and movement rates were estimated for tracked birds over a period of four years, from 2001 to 2005, in the Ecuadorian Amazon rainforest. Further details on field methods can be found in Holbrook [-@holbrook_home_2011]. We focused on the data from the many-banded aracari, *Pteroglossus pluricinctus*, because a higher number of recorded locations were available. In addition to this, the home range of this species was found to be the largest for all the toucan species studied, and therefore it was better suited to study long-distance seed dispersal events.

### Animal movement probability distributions  
What data am I using for this parameterization? It is radiotracking data, and the information on continuous trajectories for each tracked bird, to scale those distances into number of meters moved per minute for each tracking segment. I consider a tracking segment as the distance moved between two continuous radio tracking detections. (Check Kimberly's paper), because here we are taking advantage of all the tracking data since there are different tracking interval times. Then, we used that data to fit the four models described above, and did it at the different organizational levels. We considered population level data, as all the data pooled from all individuals tracked, and fit the four distributions to that data to estimate parameters. At the individual level, we fit the four pdfs for the data in respect to each individual. For the family level data, we used only a subset of the total data set to include only individuals for which family level information existed. Because these are social individuals, they tend to move with their group and so we their movements may be correlated. We wanted to explore if there were any differences at the family group levels. We fit the four pdfs to the subset of this data, as the subset population level, and then fit each distribution to the data associated to each of the family groups. We used the parameters estimated from these fits to sample step sizes for our simulation models. We also evaluated the fit of each of these distributions visually through qq plots (found in appendix A), and further details on data organization and model fitting can be found in the online repository.

### Model selection? for the step size

@@

# Statistics associated to the simulations
We analyzed simulation outputs and had seed dispersal distance as our main focus. We calculated summary statistics for each of the dispersal distances, and estimated the percentage of long distance dispersal events. Using the 500 meter threshold which was determined in previous work.

@@

@@

@@

@@

@@

@@

added Oct 2020: call for the need for more complex animal movement models when running seed dispersal simulations. Even the posibility of mixed distribution models, where not only we allow for each individual or fmailiy group to have it's own model, but also use multiple distributions for each. The short movements performed by these animals may be better described by a different distribution than those performed at longer intervals.
